---
title: "Getting started with aae.hydro"
author: "Jian Yen"
date: "30/04/2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Getting started with aae.hydro}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
# display code by default
knitr::opts_chunk$set(echo = TRUE)
```

## We have access to lots of data

We have so much data. But it's hard to navigate and work with everything, especially when we want to bring together data from different places. The `aae.hydro` R package is an attempt to simplify this process. The main aim of the package is to provide an interface to publicly available streamflow data. Down the track, the goal is to link `aae.hydro` with other packages to access data collected, curated, or regularly used by AAE staff.

## Installing the package

You can install the `aae.hydro` package from GitHub. Although the package doesn't publicly share any AAE data sets, it will not be submitted to CRAN given the limited set of users. To install from GitHub, you'll need to install the `remotes` R package and use the following lines of code:

```{r install-packages, eval = FALSE}
# install the remotes package if not already installed
install.packages("remotes")

# install the aae.hydro package from GitHub
remotes::install_github("aae.stats/aae.hydro@master")
```

Once completed, you should be able to load the `aae.hydro` package with `library(aae.hydro)`.
